On Optimality of Finite Element Integration

### 2016-02-15 TOMS_Paper/assembly.tex
Before removing rewrite rules

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}
Examples will be provided in Section~\ref{sec:se-examples}.

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-6); scheduling of all other sub-expressions (step 7).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to group identical terms within the same sub-expression. \\ 
\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.

\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions.

\item For all multilinear operands in $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of Proposition~\ref{prop:multi-struct}.}

\item Build the \textit{sharing graph} $\mathbb{G} = (\mathbb{S}, \mathbb{E})$, with $S^i \in \mathbb{S}$ representing a multilinear symbol or a temporary produced at step (3). An edge $(S^i$, $S^j)$ indicates that a product $S^i S^j$ would appear if the sub-expressions including $S^i$ and $S^j$ were expanded. Let $\mathbb{T}$ be the subset of vertices with at most one incident edge, or ``terminals''.

\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
\textit{Note: this ensures the flop-decreasing nature of the transformations at step (6)}.\\
\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several kinds of information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost will be derived.}

\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}.
\item There is no cyclic path $[S, W, ..., S]$: take $S$ and apply Strategy~\ref{strategy:ii}, similarly to (B).\\
\textit{Note: this recasts the expression as $S (...W + ...)$}.\\
\textit{Note: the idea is that the product ${S}\ {W}$ cannot be ``hidden within larger factorizations''. Further details will be provided in Section~\ref{sec:proof}}.
\item $[S, ..., V, ..., S]$ is a cycle, $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$: take $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...S^1 + ...S^n)$}.
\end{enumerate}
Go back to step (2).

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

\subsubsection{Examples}
\label{sec:se-examples}
Consider again Figure~\ref{code:multi_loopnest_a}. We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$. For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step 3 has no effect. The sharing graph is $\mathbb{G} = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $\mathbb{T} = \lbrace c_i, d_i \rbrace$. Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step 3). We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may not be optimal, since $\theta$ only distinguishes monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials sharing some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum. This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (via Strategy~\ref{strategy:ii}). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (3) cannot prune the optimum from the search space.

Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. We here demonstrate why this is a step forwards towards the optimum. Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)).

Instances of rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid regardless of how the applications of Strategy~\ref{strategy:ii} are scheduled. In particular:
\begin{description}
\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
\end{description}

Rule (C) is the critical one. Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$. Since $\mathbb{T} = \emptyset$, $S$ is also the root a second sub-graph $\mathbb{C}$. The non-trivial case in when $\mathbb{C}$ is cyclic. If any vertex in $\mathbb{C}$ were factorized prior to $S$, the product $S\ W$ could eventually appear in the transformed expression. Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted.

Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}

>>>>

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\end{itemize}
Examples will be provided in Section~\ref{sec:se-examples}.

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont 
~
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to collect and partition multilinear operands into disjoint sets, $\mathbb{P} = \lbrace P^1, ..., P^{p}\rbrace$. $\mathbb{P}$ is such that all multilinear operands in $P$ share the same set of multilinear symbols $S_P$, whereas there is no sharing across different partitions. For all multilinear operands in $P \in \mathbb{P}$ such that $|P| < |S_P|$, apply Strategy~\ref{strategy:i}. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of Proposition~\ref{prop:multi-struct}.}

\item Build the \textit{sharing graph} $G= (S, E)$. Each $s \in S$ represents a multilinear symbol or a temporary produced at step (2). An edge $(s^i$, $s^j)$ indicates that a product $s^i s^j$ would appear if the sub-expressions including $s^i$ and $s^j$ were expanded.

\item Let $T \subset S$ be the subset of vertices with at most one incident edges, or ``terminals''. Partition $T$ into disjoint sets, $\mathbb{T} = \lbrace T^1, ..., T^t\rbrace$, such that the terminals in $T^i$ are either adjacent to the same vertex or disconnected. Apply Strategy~\ref{strategy:ii} factorizing the terminals in each $T^i$, provided that this leads to a reduction in operation count.\\
\textit{Note: the last check ensures the flop-decreasing nature of the transformation; the cases in which the expansion cost exceeds the code motion gain are discarded.}

\item Partition $S$ into disjoint sets, $\mathbb{S} = \lbrace S^1, ..., S^n\rbrace$, such that $S^i$ includes all instances of a given symbol $s$ in the expression. Transform $G$ by merging $\lbrace s^1, ..., s^m \rbrace \subset S^i$ into a unique vertex $s$ (taking the union of the edges), provided that factorizing $[s^1, ..., s^m]$ would not cause an increase in operation count.\\
\textit{Note: the variation in operation count is a function of (number and arity) of the products wrapping the elements in $S^i$, which impact the expansion cost.}

\item Restructure the expression based on the structure of $G$ and a set of rewrite rules $<$\textit{precondition}: \textit{action}$>$, to be applied in the order given below.
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace t^1, ..., t^n \rbrace \subset T$ adjacent to $s \in S$, $n > 1$: take $s$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $s (...t^1 + ... + ...t^n)$}.
\item Only one terminal $t$ adjacent to $s$: take an arbitrary non-terminal symbol $s^{nt}$ adjacent to $s$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $s (...t + ...s^{nt})$}. 
\item $T = \emptyset$, $C = [s, ..., v, ..., s]$ is a cycle, $v$ has highest degree in $C$, $\lbrace s^1, ..., s^n \rbrace$ adjacent to $v$: take $v$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $v (...s^1 + ... + ...s^n)$}.
\end{enumerate}
Rule (C) may expose further hits for rules (A) and (B). Rules are therefore applied cyclically until no more hits are possible.\\
\textit{Note: the last two steps only impact bilinear forms, since otherwise $E = \emptyset$.}

@@

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of flop-decreasing transformations (steps (3) and (4)) and the re-scheduling of sub-expressions within outer loops (last step) also attempt to optimize the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

\subsubsection{Examples}
\label{sec:se-examples}
Consider again Figure~\ref{code:multi_loopnest_a}. We have $\mathbb{P} = \lbrace P^0, P^1, P^2\rbrace$, with $P^0 = \lbrace b_j\rbrace$, $P^1 = \lbrace c_i\rbrace$, and $P^2 = \lbrace d_i\rbrace$. For all $P^i$, we have $|P^i| = 1 = |S_{P^i}|$, although applying Strategy~\ref{strategy:i} in step (1) has no effect. The sharing graph is $G = (\lbrace b_j, c_i, d_i \rbrace, \lbrace (b_j, c_i), (b_j, d_i) \rbrace$, and $T = \lbrace c_i, d_i \rbrace$. Rewrite rule (4A) is hit, which leads to the code in Figure~\ref{code:multi_loopnest_c}.

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so Strategy~\ref{strategy:i} is applied to both partitions (step (1)). We then have (step (3)) $G = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \lbrace (t^0, t^2), (t^1, t^3) \rbrace)$, although no rewrite rule is applicable since $S = T$

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may not be optimal, since $\theta$ only distinguishes monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials sharing some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a non-optimum. This situation is prevented by construction, because $\Gamma$ exhaustively searches all possible bipartitions for determining an optimum preserving Constraint~\ref{const:TH}\footnote{Note that the problem can be seen as an instance of the well-known Knapsack problem}. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Proposition~\ref{prop:multi-struct} ensures that there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through generalized code motion (Strategy~\ref{strategy:i}) or factorization of multilinear symbols (via Strategy~\ref{strategy:ii}). If applied, these two strategies would lead, respectively, to performing $|P|$ and $|S_P|$ multiplications at every loop iteration. Since Strategy~\ref{strategy:i} is applied if and only if $|P| < |S_P|$ and does not change the structure of the expression (it requires neither expansion nor factorization), step (1) cannot prune the optimum from the search space.

After structuring the sharing graph $G$ in such a way that only flop-decreasing transformations are possible, a set of rewrite rules are applied. We here discuss why applying, in rigorous order, rules (A), (B), and (C), leads to the optimum.

Rules (A) and (B) represent ``unavoidable'' factorizations, for which one multiplication must be paid no matter how Strategy~\ref{strategy:ii} is scheduled. In particular:
\begin{description}
\item[Rule (A)] The terminals $\lbrace t^1, ..., t^n \rbrace$ can only be grouped through the factorization of $s$ (by definition of terminal), so this rule has no side effects.
\item[Rule (B)] Since $s^{nt}$ is not a terminal, there must exist at least one edge between $s^{nt}$ and a generic vertex $v$. Let us consider the alternative factorization $s^{nt}(...s + ... + ...v)$. If $v$ were a terminal, we observe that setting $v = t$ and $s = s^{nt}$ would simply expose another hit for rule (B). Otherwise ($v$ non terminal), we would have $s$ now appearing in two different sub-expressions ($(...s + ... + ...v)$ and $(...s\ t)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $s$. From this, we evince that applying (C) prior (B) may prune the optimum from the search space.
\end{description}
To reach the optimum, rule (C) discards all those potential factorizations that can be rather ``aggregated'' with \textit{essential} factorizations. A factorization is essential if a vertex is in a cycle. Being in a cycle is the fundamental condition to

Having higher degree simply means grouping more symbols (i.e., avoiding more products);

%Step (6) applies factorization following the ordering dictated by the sharing graph and its rewrite rules. We here demonstrate why this is a step forwards towards the optimum. Note that, at this point, non flop-decreasing factorizations have already been pruned (step (5)). 
%
%Instances of 
%\begin{description}
%\item[Rule (A)] As terminals, $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule has no side effects.
%\item[Rule (B)] Since $S^{nt}$ is not a terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ were a terminal, we see that setting $V = T$ and $S = S^{nt}$ would simply expose another hit for rule (B). Otherwise ($V$ non terminal), we would have $S$ now appearing in two different sub-expressions ($(...S + ... + ...V)$ and $(...S\ T)$), which would clearly be sub-optimal as a consequence of the spatial sharing induced by $S$. From this, we evince that applying rules (C) and (D) prior to rule (B) may prune the optimum from the search space.
%\end{description}
%
%Rule (C) is the critical one. Let $\mathbb{A}$ be an acyclic sub-graph rooted in $S$, and let $W$ be the neighbor of $S$ in $\mathbb{A}$. Since $\mathbb{T} = \emptyset$, $S$ is also the root a second sub-graph $\mathbb{C}$. The non-trivial case in when $\mathbb{C}$ is cyclic. If any vertex in $\mathbb{C}$ were factorized prior to $S$, the product $S\ W$ could eventually appear in the transformed expression. Conversely, is Strategy~\ref{strategy:ii} were first applied to $S$, $W$ and one or more symbols in $\mathbb{C}$ would be grouped in the same sub-expression and hoisted. 
% 
%Finally, note that the algorithm is guaranteed to terminate because of the jump to step (7) as $\mathbb{S} = \mathbb{T}$. 
\end{enumerate}
\end{proof}
